### 2020-03-10 Common/chapters/medical.tex
Julien and Johan: minors

\section{NIfTI file format}
\subsection{Origin}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}The Neuroimaging Informatics Technology Initiative (NIfTI) file format is the successor of the ANALYZE file format. The main problem of the latter was that it was lacking information about orientation in space. Therefore, the interpretation of stored data could be problematic and inconsistent. For instance, there was a real confusion to determine the left and right sides of brain images. Hence, the NIfTI file format was defined to overcome this major issue.

@@

\subsection{DICOM}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}Visualizing DICOM files is pretty straightforward since each file represents a single two-dimensional image. However, a lot of pixel transformations and normalizations have to be applied to obtain the desired result (see Section \ref{sec:dicom_data_manipulation}), which may be a source of errors. Our tool allows to display a single DICOM file as well as a sequence of files if the function is fed with a directory. Users can then scroll through the Z-axis, displaying the next or previous slice.  Figure \ref{fig:visualize_lung_dcm} shows the visualization of a lung at three different Z-axis coordinates. This allows the user to have a complete overview of the body part he wants to analyze.

@@

\subsection{RAW}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}Medical RAW files are three-dimensional, which means that a single file contains a volume, i.e. a succession of 2D slices. Scrolling with the mouse goes through slices over a specific axis. Like the NIfTI file format, the user can profit from the three-dimensionality by displaying the body part under three different perspectives. This tool was particularly useful since no free software capable of handling these files was available. Photography-related programs such as Photoshop can open RAW pictures but not these three-dimensional ones.

@@

\section{Conversion to PNG}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}In addition to the visualization tools, libraries allowing to convert medical files into PNG files were developed. The main one is called "anydatasettopng" and is the one to use to convert any file format to PNG as it makes use of the others. Given a root directory, it scans the latter and its subdirectories to find all DICOM, NIfTI and RAW/MHD files. Then, each file is converted to PNG and saved into the same directory as the one it came from. This library allows to keep an easily accessible visual representation of medical files without having to rely on a specific piece of software to display them. Finally, it also played the role of a debugging tool since the conversion techniques are the same as the ones used further in this work.

>>>>

\section{NIfTI file format}
\subsection{Origin}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}The Neuroimaging Informatics Technology Initiative (NIfTI) file format is the successor of the ANALYZE file format. The main problem of the latter was a lack of information about orientation in space. Therefore, the interpretation of stored data could be problematic and inconsistent. For instance, there was a real confusion to determine the left and right sides of brain images. Hence, the NIfTI file format was defined to overcome this major issue.

@@

\subsection{DICOM}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}Visualizing DICOM files is pretty straightforward since each file represents a single two-dimensional image. However, a lot of pixel transformations and normalizations have to be applied to obtain the desired result (see Section \ref{sec:dicom_data_manipulation}), which may be a source of errors. Our tool allows to display a single DICOM file as well as a sequence of files if the function is fed with a directory. Users can then scroll through the z-axis, displaying the next or previous slice.  Figure \ref{fig:visualize_lung_dcm} shows the visualization of a lung at three different z-axis coordinates. This allows the user to have a complete overview of the body part he wants to analyze.

@@

\subsection{RAW}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}Medical RAW files are three-dimensional, which means that a single file contains a volume, i.e. a succession of 2D slices. Scrolling with the mouse goes through slices over a specific axis. Like the NIfTI file format, the user can profit from the three-dimensionality by displaying the body part under three different perspectives. This tool is particularly useful as no free software capable of handling these files is available. Photography-related programs such as Photoshop can open RAW pictures but not these three-dimensional ones.

@@

\section{Conversion to PNG}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Cl{\'e}ment}}In addition to the visualization tools, scripts allowing to convert medical files into PNG files were developed. The main one is called "anydatasettopng" and is the one to use to convert any file format to PNG as it makes use of the others. Given a root directory, it scans the latter and its subdirectories to find all DICOM, NIfTI and RAW/MHD files. Then, each file is converted to PNG and saved into the same directory as the one it came from. This script allows to keep an easily accessible visual representation of medical files without having to rely on a specific piece of software to display them. Finally, it also played the role of a debugging tool since the conversion techniques are the same as the ones used further in this work.

### 2020-03-10 Common/chapters/paper_reproduction.tex
Julien and Johan: minors

Song et al.~\cite{07} proposed a deep convolutional neural network (DCNN) method to detect prostate cancer based on the SPIE-AAPM-NCI PROSTATEx Challenge dataset. This dataset is one of the biggest datasets available for prostate cancer classification. The two different output classes of the latter are benign lesion (class 0) and malignant lesion (class 1). The dataset contains a total of 204 patients for the training (labels are known) and 208 patients for the challenge (labels are unknown). This paper explains all the steps to follow with the goal of building a computer-aided diagnosis system for prostate cancer detection. Moreover, it also provides results (AUC) about the algorithm performance on a test set they built themselves (which is not the same as the challenge test set). In fact, they split the PROSTATEx training set into a training set, a validation set and the mentioned test set. These results can be used as a benchmark to compare with the results of the reproduction of the experiment. However, the article gives no information about the results their model achieved on the official PROSTATEx challenge test set (208 patients without labels). Therefore, the reproduction of the experiment will fill that gap by taking part in the actual challenge.

@@

\section{Process overview}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Schematically speaking, the entire experiment process can be represented as shown on Figure \ref{fig:paper_reproduction_process}. The first part of the experiment is about the reproduction of the paper experiment, whereas the second one is devoted to the participation in the SPIE-AAPM-NCI PRO\- STATEx challenge since the authors of the article did not take part in it.

\subsubsection{Experiment reproduction}
The first part of the experiment makes use of the PROSTATEx training set. This dataset is only composed of samples whose clinical significance is provided (labeled data). The first step of the reproduction is the data processing. In fact, many processing steps are involved to transform the original DICOM files into NumPy arrays which can be fed to the neural network. Data processing includes lesion cropping, normalization, MRI images stacking, data augmentation, etc. (see Section \ref{sec:prostatex_data_processing} for further details).\\
Once the processing part is completed, the labeled data are split into a training set, a validation set and a test set using respectively 80\%, 10\% and 10\% of the available data for each subset. The training set samples are then fed to the neural network in batches. 
%, which updates its weight consequently. 
At the end of each epoch, the current model is tested on the validation set and the resulting metrics are plotted in Tensorboard (see Section \ref{paper_tensorboard}). The validation set is independent of the training set and has never been seen by the model, which gives an indication of how well the latter generalizes. Then, if the current model reached a higher AUC and a higher accuracy than the previous best model, it becomes the new best model and it is saved. Furthermore, if the validation AUC decreases for a defined number of epochs, the learning rate is reduced to prevent overfitting and avoid a progressive decrease in validation performance. This technique contributes to maintaining a good generalization level and optimizes the training with respect to the AUC.\\
As soon as the training phase reaches the defined number of epochs, the model is tested on the test set with the authors' method called "enhanced prediction" to compute the AUC. The latter method consists in predicting an output for each augmented test image before averaging the predictions for each lesion of each patient.

@@

Once full images were converted to NumPy arrays, the images related to a specific lesion of a specific patient needed to be aligned and resampled to the same resolution. In fact, T2-weighted images have a much higher resolution than the DWI and ADC images (in this dataset at least). Without this operation, a single pixel on a DWI image would have covered a lot more tissue than a pixel of the corresponding T2 image. To perform the alignment, the lesion was localized on the three sequences than to the CSV file. Then, the goal was to crop a large patch which contained the same amount of tissue on the three images. This patch had to be centered on the lesion to ease the augmentation process. So, a fixed patch size was defined for the T2 image since its resolution was the highest. Then, using the voxel spacing information, the patch size required to cover the same amount of tissue was computed for the two other sequences (DWI, ADC). Finally, the three images were cropped, resized to the same resolution and stacked. Stacking images consists in putting each single image (grayscale) into a single array, as three different channels. At this point, the pixel $i,j$ of the three channels represents the exact same tissue, which means that the lesion is at the same position and is covered by the same number of pixels on each channel. This approach increases the probability of detecting a cancer by ensuring a good visibilty for each lesion, since the latter is not necessarily as visible on the three images.

@@

This way of proceeding gave awful results. The main reason is that images from different sequences show the same tissue in an almost opposite way. The most striking difference to the human eye is the colors. Tumors are usually white on DWI images and black on ADC images. Moreover, they are not always as visible on the images of the various sequences. Therefore, finding similar features on all sequences was extremely difficult. Stacking the three sequences solves this problem. 
%In this case, the first 4225 input neurons (4225 input neurons for a $65$x$65$px image) are dedicated to T2 images, the next 4225 neurons to DWI images and the last 4225 neurons to ADC images.

@@

\subsection{Discussion}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Thanks to the curves of Section \ref{sec:paper_reproduction_results}, a lot of information regarding the model performance can be inferred. In fact, "reviewing learning curves of models during training can be used to diagnose problems with learning, such as an underfit or overfit model, as well as whether the training and validation datasets are suitably representative"~\cite{40}.

First of all, when observing the loss plot, what is noticeable is that the training and validation losses have the same behavior during the 50 epochs. This is a clear sign of good fit of the model on the data. All the features learned by the model on the training data are also applicable to the validation set. This situation is the optimal one. Regarding the slope of the curve, a big change occurs at epoch 17. This moment corresponds to the one where the learning rate is reduced by a factor of $100$ thanks to the \textit{reducelronplateau} function. In fact, at this moment, the validation AUC (the training is optimized according to this metric) was at its optimum with this learning rate. If this technique had not been used, the training loss would have continued to go down whereas the validation loss would have gone up immediately, which would have led to overfitting. Here, thanks to the learning rate scheduler, this scenario is avoided as the validation and training losses stay stable from this point onwards. The same behaviour is clearly noticeable for all other metrics.

Regarding the overall performance of the best model chosen at epoch 21, the latter generalizes well. This is particularly visible on the accuracy and AUC plots where the validation metrics are higher or equal to the training ones. The validation accuracy indicates that the model made the right class decision in approximately $75$\% of the time. Furthermore, since the model achieved a validation AUC of $0.76$, it shows that it had a good distinction capacity. In fact, the highest the AUC is, the highest is the ability of the model to predict benign lesions as benign and malignant lesions as malignant. Armato et al.~\cite{42} claim that a less-experienced radiologist can reach an AUC of $0.81$ and an expert an AUC of $0.91$. Consequently, this model is almost at the level of a radiologist, which is very promising. It is also important to keep in mind that the model was trained using only 164 patients (80\% of the 204 training patients), which suggests that even better results can be reached with more data.

@@

\noindent Another extremely important aspect of a computer-aided diagnosis system is to avoid the number of false negatives as much as possible. In fact, the consequences of classifying a malignant lesion as benign are far worse than the opposite. This concept is measured by the recall, which quantifies the proportion of malignant lesions correctly classified among all malignant lesion. The model has a recall of around $0.74$ on the validation set.

@@

First, in their work, the cropping of lesions was performed by a qualified radiologist, whereas, in our case, it was automatically done using the information in the CSV files. Then, the way the test set is built can make the results vary a lot. In fact, the authors did not provide any information about how they built their test set. It is therefore possible that the test set was chosen to maximize the performance of the model. In the current work, the test set is built randomly. Consequently, the comparison of the performance is difficult since the test set is not the same in the two works. The only way to seriously compare the two models would be to compare their performance on the official SPIE-AAPM-NCI Prostate MR classification challenge test set. The model of this section (trained with 80\% of all the data available) reached an AUC of $0.71$. Unfortunately, no information about the performance of the model of the paper is given. Knowing that the best result at the time of the challenge was an AUC of $0.87$ on the challenge test set, it is highly surprising not to take part in it when claiming an AUC in the interval $[0.876-0.994]$ with 95\% of confidence. Finally, another source of difference is the number of participants in the experiment and the level of experience they have. Indeed, seven authors contributed to the paper, including one student with a Bachelor degree and six PhD with one of them working for Siemens Healthcare and one for hImagingTek Ltd. These big companies are specialized in medical imaging and have departments dedicated to machine learning. On the opposite, this work was produced independently by two MSc students without previous experience in the medical imaging field.

>>>>

Song et al.~\cite{07} proposed a deep convolutional neural network (DCNN) method to detect prostate cancer based on the SPIE-AAPM-NCI PROSTATEx Challenge dataset. This dataset is one of the biggest datasets available for prostate cancer classification. The two different output classes of the latter are benign lesion (class 0) and malignant lesion (class 1). The dataset contains a total of 204 patients for the training (labels are known) and 208 patients for the challenge (labels are unknown). This paper explains all the steps to follow with the goal of building a deep learning model for prostate cancer detection. Moreover, it also provides results (AUC) about the algorithm performance on a test set they built themselves (which is not the same as the challenge test set). In fact, they split the PROSTATEx training set into a training set, a validation set and the mentioned test set. These results can be used as a benchmark to compare with the results of the reproduction of the experiment. However, the article gives no information about the results their model achieved on the official PROSTATEx challenge test set (208 patients without labels). Therefore, the reproduction of the experiment will fill that gap by taking part in the actual challenge.

@@

\section{Process overview}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Schematically speaking, the entire experiment process can be represented as shown on Figure \ref{fig:paper_reproduction_process}. The first part of the experiment is about the reproduction of the paper experiment, whereas the second one is devoted to the participation in the SPIE-AAPM-NCI PROSTATEx challenge since the authors of the article did not take part in it.

\subsubsection{Experiment reproduction}
The first part of the experiment makes use of the PROSTATEx training set. This dataset is only composed of samples whose clinical significance is provided (labeled data). The first step of the reproduction is the data processing. In fact, many processing steps are involved to transform the original DICOM files into NumPy arrays which can be fed to the neural network. Data processing includes lesion cropping, normalization, MRI images stacking, data augmentation, etc. (see Section \ref{sec:prostatex_data_processing} for further details).\\
Once the processing part is completed, the labeled data are split into a training set, a validation set and a test set using respectively 80\%, 10\% and 10\% of the available data for each subset. The training set samples are then fed to the neural network in batches. 
%, which updates its weight consequently. 
At the end of each epoch, the current model is tested on the validation set and the resulting metrics are plotted in Tensorboard (see Section \ref{paper_tensorboard}). The validation set is independent of the training set and has never been seen by the model, which gives an indication of how well the latter generalizes. Then, if the current model reaches a higher AUC and a higher accuracy than the previous best model, it becomes the new best model and it is saved. Furthermore, if the validation AUC decreases for a defined number of epochs, the learning rate is reduced to prevent overfitting and avoid a progressive decrease in validation performance. This technique contributes to maintaining a good generalization level and optimizes the training with respect to the AUC.\\
As soon as the training phase reaches the defined number of epochs, the model is tested on the test set with the authors' method called "enhanced prediction" to compute the AUC. The latter method consists in predicting an output for each augmented test image before averaging the predictions for each lesion of each patient.

@@

Once full images were converted to NumPy arrays, the images related to a specific lesion of a specific patient needed to be aligned and resampled to the same resolution. In fact, T2-weighted images have a much higher resolution than the DWI and ADC images (in this dataset at least). Without this operation, a single pixel on a DWI image would have covered a lot more tissue than a pixel of the corresponding T2 image. To perform the alignment, the lesion was localized on the three sequences than to the CSV file. Then, the goal was to crop a large patch which contained the same amount of tissue on the three images. This patch had to be centered on the lesion to ease the augmentation process. So, a fixed patch size was defined for the T2 image since its resolution was the highest. Then, using the voxel spacing information, the patch size required to cover the same amount of tissue was computed for the two other sequences (DWI, ADC). Finally, the three images were cropped, resized to the same resolution and stacked. Stacking images consists in putting each single image (grayscale) into a single array, as three different channels. At this point, the pixel $i,j$ of the three channels represents the exact same tissue, which means that the lesion is at the same position and is covered by the same number of pixels on each channel. This approach increases the probability of detecting a cancer by ensuring a good visibilty for each lesion, as the latter is not necessarily as visible on the three images.

@@

This way of proceeding gave awful results. The main reason is that images from different sequences show the same tissue in an almost opposite way. The most striking difference to the human eye is the colors. Tumors are usually white on DWI images and black on ADC images. Moreover, they are not always as visible on the images of the various sequences. Therefore, finding similar features on all sequences is extremely difficult. Stacking the three sequences solves this problem. 
%In this case, the first 4225 input neurons (4225 input neurons for a $65$x$65$px image) are dedicated to T2 images, the next 4225 neurons to DWI images and the last 4225 neurons to ADC images.

@@

\subsection{Discussion}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin}}Thanks to the curves of Section \ref{sec:paper_reproduction_results}, a lot of information regarding the model performance can be inferred. In fact, as stated by Jason Brownlee, "reviewing learning curves of models during training can be used to diagnose problems with learning, such as an underfit or overfit model, as well as whether the training and validation datasets are suitably representative"~\cite{40}.

First of all, when observing the loss plot, it is noticeable that the training and validation losses have the same behavior during the 50 epochs. This is a clear sign of good fit of the model on the data. All the features learned by the model on the training data are also applicable to the validation set. This situation is the optimal one. Regarding the slope of the curve, a big change occurs at epoch 17. This moment corresponds to the one where the learning rate is reduced by a factor of $100$ thanks to the \textit{reducelronplateau} function. In fact, at this moment, the validation AUC (the training is optimized according to this metric) was at its optimum with this learning rate. If this technique had not been used, the training loss would have continued to go down whereas the validation loss would have gone up immediately, which would have led to overfitting. Here, thanks to the learning rate scheduler, this scenario is avoided as the validation and training losses stay stable from this point onwards. The same behaviour is clearly noticeable for all other metrics.

Regarding the overall performance of the best model chosen at epoch 21, the latter generalizes well. This is particularly visible on the accuracy and AUC plots where the validation metrics are higher or equal to the training ones. The validation accuracy indicates that the model made the right class decision in approximately $75$\% of the time. Furthermore, as the model achieves a validation AUC of $0.76$, it shows that it has a good distinction capacity. In fact, the highest the AUC is, the highest is the ability of the model to predict benign lesions as benign and malignant lesions as malignant. Armato et al.~\cite{42} claim that a less-experienced radiologist can reach an AUC of $0.81$ and an expert an AUC of $0.91$. Consequently, this model is almost at the level of a radiologist, which is very promising. It is also important to keep in mind that the model was trained using only 164 patients (80\% of the 204 training patients), which suggests that even better results can be reached with more data.

@@

\noindent Another extremely important aspect in the medical field is to avoid the number of false negatives as much as possible. In fact, the consequences of classifying a malignant lesion as benign are far worse than the opposite. This concept is measured by the recall, which quantifies the proportion of malignant lesions correctly classified among all malignant lesion. The model has a recall of around $0.74$ on the validation set.

@@

First, in their work, the lesion cropping was performed by a qualified radiologist, whereas, in our case, it was automatically done using the information in the CSV files. Then, the way the test set is built can make the results vary a lot. In fact, the authors did not provide any information about how they built their test set. It is therefore possible that the test set was chosen to maximize the performance of the model. In the current work, the test set is built randomly. Consequently, the comparison of the performance is difficult since the test set is not the same in the two works. The only way to seriously compare the two models would be to compare their performance on the official SPIE-AAPM-NCI Prostate MR classification challenge test set. The model of this section (trained with 80\% of all the data available) reached an AUC of $0.71$. Unfortunately, no information about the performance of the model of the paper is given. Knowing that the best result at the time of the challenge was an AUC of $0.87$ on the challenge test set, it is highly surprising not to take part in it when claiming an AUC in the interval $[0.876-0.994]$ with 95\% of confidence. Finally, another source of difference is the number of participants in the experiment and the level of experience they have. Indeed, seven authors contributed to the paper, including one student with a Bachelor degree and six PhD with one of them working for Siemens Healthcare and one for hImagingTek Ltd. These big companies are specialized in medical imaging and have departments dedicated to machine learning. On the opposite, this work was produced independently by two MSc students without previous experience in the medical imaging field.

### 2020-03-10 Common/chapters/transfer_learning.tex
Julien and Johan: minors

\section{Goal}
\label{sec:tl_goal}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin \& Cl{\'e}ment}}Transfer learning consists in using a network which was pretrained on a dataset A to improve the performance of the task on a dataset B.

@@

In this experiment, the goal is to increase the model performance on the test set, particularly the AUC. In fact, the latter represents a degree of separability, i.e. how well the model distinguishes between classes. In cancer classification, the diagnosis must be clear. For instance, Model A and Model B can classify a sample X as benign (prediction: class 0) but with different probabilities of 0.10 and 0.45 respectively, where the latter are the probabilities of belonging to class 1 (malignant). While Model A is sure of choice, Model B is not trustworthy and may misdiagnose samples.

>>>>

\section{Goal}
\label{sec:tl_goal}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Jobin \& Cl{\'e}ment}}Transfer learning consists in using a network which was pretrained on a dataset A to improve the performance of a similar task on a dataset B.

@@

Throughout the whole pipeline, the target metric to optimize is the AUC. In fact, the latter represents a degree of separability, i.e. how well the model distinguishes between classes. In cancer classification, the diagnosis must be clear. For instance, Model A and Model B can classify a sample X as benign (prediction: class 0) but with different probabilities of 0.10 and 0.45 respectively, where the latter are the probabilities of belonging to class 1 (malignant). While Model A is sure of its choice, Model B is not trustworthy and may misdiagnose samples.
